Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-09 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.tex
edit latex document for reverse-inclusion into rmarkdown document

One of few methodologies to address all three of these established needs
is \hl{similarity cohort--based modeling (SCBM)}, which
\hl{we find in this review} to have been introduced several times
independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that
relies on a \hl{patient similarity measure }\citep{Brown2016} to extract
a cohort of past or training cases that then inform the diagnosis,
prognosis, or care of a new or test case.\footnote{Similarity measures
  between more granular units of analysis, such as encounters and
  decision points\hl{, or between patients' time series}, are often
  still referred to as patient similarity measures\hl{ in our sample}.}
\hl{Some of the earliest and simplest CBR tools returned} these
retrieved cases \hl{ to inform human decision-making}\citep{Aamodt1994};
the related nearest neighbors (NN) technique generates predictions from
cohorts automatically via averaging (regression) or voting
(classification) of their outcomes. In contrast, \hl{SCBM} fits families
of predictive models, for example generalized linear models (GLMs), to
retrieved similarity cohorts in order to generate predictions.
\hl{While many ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions }\citep{Molnar2023}\hl{, these quantifications are not directly interpretable model components analogous to the split nodes of a DT or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected.}

\hl{This approach offers an alternative paradigm for interpretable machine learning that combines the simplicity and transparency of classical models like decision trees (DTs) and GLMs with the sensitivity to nonlinearity and interactions of advanced opaque models like random forests (RFs) and artificial neural networks (ANNs). The sensitivity is enabled by the fitting of the model to the local cohort, whose size and scope are governed by hyperparameters that may be tuned like any others via grid or Bayesian optimization. Some similarity measures or other retrieval methods may be more interpretable than others, so that the interpretability of the SCBM approach would then depend on that of its retrieval (similarity matching) and revision (predictive modeling) steps. Moreover, this approach would benefit computationally as well as practically from the availability of a large corpus of cases, as efficient data structures and algorithms exist for similarity matching }\citep{Halder2024}\hl{ and the local models would benefit from increased numbers of relevant cases.}
\hl{SCBM}s thus provide an individualized way to fit fully interpretable
models to large population data.

@@

\hl{We also note that a large variety of AI tools, besides NN, use similarity measures or other relevance-based retrieval methods: Back-propagation in ANNs relies on differentiation, which requires that all network nodes' activation functions and, therefore, the final model are continuous and differentiable, i.e. that small distances between cases yield small changes in prediction; and kernel-based methods that rely directly on pairwise distances computed via inner product, rather than high-dimensional coordinates, extend far beyond their use in CBR. The key difference in SCBM is that the retrieved cohort is treated as a component of the model and made available to the user. Indeed, retrieval procedures were the most significant algorithmic requirements and accomplishments of CBR }\citep{Kolodner1992, Gierl1998}\hl{ and the retrieved cohorts may be reused for numerous modeling tasks.}
Thus, \hl{SCBM}s may embody both styles of CBR.

@@

Figure \ref{fig:prisma} depicts our identification of studies via
databases and registers. Following the completion and input of each
search string, there were a total of 1,817 sources within all of the
platforms used. De-duplication resulted in 1,657 entries, which were
added to the title/abstract screening for review. Of these, 360 entries
met the screening criteria and were assessed through full text review.
Of these, 51 fit the original criteria, and 21 were included as distinct
from NN prediction.

@@

We then reviewed \hl{a seed set comprising 43 studies} that inspired
this review. After removing duplicates and screening for eligibility, we
were left with 6 additional studies
\citep{Park2006, Lowsky2013, Lee2015, Ng2015, Lee2017, Wang2019}, 1 of
which was excluded from the synthesis for using NN prediction. Reference
tracking from the 43 + 360 = 403 studies assessed for eligibility led us
to identify 1 additional study that met our criteria
\citep{Kasabov2010}. This left us with 21 + 1 + 5 = 27 studies included
in the review and synthesis.

@@

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While these studies trace back to the late 1990s, they
have become \hl{slightly }more common, which suggests that this is an
active, though not rapidly expanding, approach. We note that CBR in
health and medicine originated as early as 1990, and \hl{SCBM} emerged
soon after CBR had established itself; the idea has been ``in the air''
for as long as CBR has been in use.

@@

While all included studies were reported in scientific and medical
journals, the vast majority were oriented toward clinical practice
rather than medical research. For example, a 2006 study specifically
evaluated the usefulness of CBR-based explanations for the purpose of
decision support \citep{Doyle2006}. This study was part of a much larger
literature on CBR systems and was included here despite relying on NN
prediction because it used an unconventional voting scheme to generate
recommendations. A more recent study took essentially the same focus
with respect to a proposed clinical risk prediction model, which
amounted to CBR with a novel weighting scheme on predictors informed by
expert consensus \citep{Fang2021}. In both cases a prototype
implementation was \hl{evaluated} in an experimental \hl{setting.}

@@

\hl{Eight} studies used response values in training data sets to
supervise similarity learning while two used unsupervised learning
(Table \ref{tab:composite}). Based on the key properties of DML
algorithms identified by \citet{Bellet2014}, most learned measures were
non-linear while some were locally learned, and some were optimized
globally while others locally. Not all qualified as DML, since the
resulting measures would not necessarily satisfy the triangle
inequality. Each used its similarity measure to retrieve training cases
relevant to each testing case. \hl{Most} took a standard NN approach to
generating predictions; \hl{two} \citep{Vilhena2016, Tang2021} fit
predictive models to the retrieved cohorts.

@@

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by \citet{Park2006} to
optimize the predictive performance of models fitted to cohorts
retrieved using a fixed similarity rather than count threshold, a
counterpart to conventional CBR they called ``statistical
CBR''.\footnote{Drawing from \citet{Goyal2008}, we suggest the term
  ``combinatorial CBR'' for the then-conventional approach using
  similarity cohorts of fixed cardinality.} The step used a heuristic
procedure to locate a similarity threshold that (locally) maximizes
predictive accuracy on the training set. This is analogous to optimizing
the neighborhood size parameter in NN predictive modeling, so we will
not discuss it further except to mention that the authors found
statistical CBR to outperform conventional CBR on several data sets.
\citet{CampilloGimenez2013} combined these learning techniques: They
used logistic regression on the training set to obtain weights for the
predictors (by predicting the binary outcome of kidney transplant
waitlist registration) and for the cases (by predicting agreement of
outcome between an index case and other training cases), and they used
exhaustion to optimize the size of the retrieved cohort for the accuracy
of the NN model using the previously optimized weights.

@@

\hl{One theme} was the optimization of retrieved cohort sizes for some
measure of model performance. Despite the early recommendation by
\citet{Park2006} to threshold cohorts by similarity rather than by
cardinality, only the one previously published study
\citep{Mariuzzi1997} and the most recent study \citep{Tang2021, Ng2021}
in this corpus took this approach, though \citet{Liu2022} took a third
option by retrieving fixed proportions of training data. Viewed as a
hyperparameter of the individualized modeling approach, cohort size can
be treated in the same way as neighborhood size in NN \hl{prediction},
so we will not discuss it further.

@@

\hl{While most studies identified outstanding technical needs (Section }\ref{identified-needs}\hl{), few (though some) emphasized the need to assess human-focused qualities like user interface and user experience, interpretability, meaningfulness, or trust.
Most studies identified interpretability as an advantage of localized models, though none evaluated model interpretability and few proposed new interpretative uses.
Lack of trust in ML tools is a long-recognized problem that direct intepretability of model components could help alleviate.
One valuable direction for future work would be to measure the utility of interpretable components in research and in clinical practice and the correctness and confidence of users in their interpretations.
Another would be to compare the localized predictor importance measures obtained from localized models to the local importance measures used to explain predictions made by opaque models.}

@@

\begin{itemize}
\tightlist
\item
  CBR: case-based reasoning
\item
  CIS: clinical information system
\item
  \hl{CDSS:} clinical decision support\hl{ system}
\item
  AI: artificial intelligence
\item
  NN: nearest neighbors
\item
  GLM: generalized linear model
\item
  \hl{DT: decision tree}
\item
  \hl{RF: random forest}
\item
  \hl{ANN: artificial neural network}
\item
  ML: machine learning
\item
  PMC: PubMed Central
\item
  LOOCV: leave-one-out cross-validation
\item
  GSA: gravitational search algorithm
\item
  DML: distance metric learning
\end{itemize}

@@

& & & & \hspace{6em} & NN & 88.29\%\hspace{6em} \\

@@

& & & & \hspace{6em} & NN & 82.22\%\hspace{6em} \\

@@

& & Breast & Statistical CBR & 96.96\%\hspace{6em} & NN &

@@

& & & & \hspace{6em} & NN & 65.39\%\hspace{6em} \\

@@

& & & & \hspace{6em} & NN & 55.59\%\hspace{6em} \\

@@

\citet{Malykh2018} & Accuracy & J13/pneumonia & Case-based NN &

@@

& & K80.1/gallbladder & Case-based NN & 76.7\%\hspace{6em} & &

@@

& & H25.1/age-related cataract & Case-based NN & 94.9\%\hspace{6em} & &

@@

& & H26.2/complicated cataract & Case-based NN & 91.4\%\hspace{6em} & &

@@

& & 167.4/encephalopathy & Case-based NN & 72.4\%\hspace{6em} & &

@@

& & 167.9/CBD & Case-based NN & 75.4\%\hspace{6em} & & \hspace{6em} \\
& & N20.1/ureter & Case-based NN & 58.7\%\hspace{6em} & &

>>>>

One of few methodologies to address all three of these established needs
is \hl{similarity cohort--based modeling (SCBM)}, which
\hl{we find in this review} to have been introduced several times
independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that
relies on a \hl{patient similarity measure }\citep{Brown2016} to extract
a cohort of past or training cases that then inform the diagnosis,
prognosis, or care of a new or test case.\footnote{Similarity measures
  between more granular units of analysis, such as encounters and
  decision points\hl{, or between patients' time series}, are often
  still referred to as patient similarity measures\hl{ in our sample}.}
\hl{Some of the earliest and simplest CBR tools returned} these
retrieved cases \hl{ to inform human decision-making}\citep{Aamodt1994};
the related nearest neighbors (\hl{kNN}) technique generates predictions from
cohorts automatically via averaging (regression) or voting
(classification) of their outcomes. In contrast, \hl{SCBM} fits families
of predictive models, for example generalized linear models (GLMs), to
retrieved similarity cohorts in order to generate predictions.
\hl{While many ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions }\citep{Molnar2023}\hl{, these quantifications are not directly interpretable model components analogous to the split nodes of a DT or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected.}

\hl{This approach offers an alternative paradigm for interpretable machine learning that combines the simplicity and transparency of classical models like decision trees (DTs) and GLMs with the sensitivity to nonlinearity and interactions of advanced opaque models like random forests (RFs) and artificial neural networks (ANNs). The sensitivity is enabled by the fitting of the model to the local cohort, whose size and scope are governed by hyperparameters that may be tuned like any others via grid or Bayesian optimization. Some similarity measures or other retrieval methods may be more interpretable than others, so that the interpretability of the SCBM approach would then depend on that of its retrieval (similarity matching) and revision (predictive modeling) steps. Moreover, this approach would benefit computationally as well as practically from the availability of a large corpus of cases, as efficient data structures and algorithms exist for similarity matching }\citep{Halder2024}\hl{ and the local models would benefit from increased numbers of relevant cases.}

@@

\hl{We also note that a large variety of AI tools, besides kNN, use similarity measures or other relevance-based retrieval methods: Back-propagation in ANNs relies on differentiation, which requires that all network nodes' activation functions and, therefore, the final model are continuous and differentiable, i.e. that small distances between cases yield small changes in prediction; and kernel-based methods that rely directly on pairwise distances computed via inner product, rather than high-dimensional coordinates, extend far beyond their use in CBR. The key difference in SCBM is that the retrieved cohort is treated as a component of the model and made available to the user. Indeed, retrieval procedures were the most significant algorithmic requirements and accomplishments of CBR }\citep{Kolodner1992, Gierl1998}\hl{ and the retrieved cohorts may be reused for numerous modeling tasks.}
Thus, \hl{SCBM}s may embody both styles of CBR.

@@

Figure \ref{fig:prisma} depicts our identification of studies via
databases and registers. Following the completion and input of each
search string, there were a total of 1,817 sources within all of the
platforms used. De-duplication resulted in 1,657 entries, which were
added to the title/abstract screening for review. Of these, 360 entries
met the screening criteria and were assessed through full text review.
Of these, 51 fit the original criteria, and 21 were included as distinct
from \hl{kNN} prediction.

@@

We then reviewed \hl{a seed set comprising 43 studies} that inspired
this review. After removing duplicates and screening for eligibility, we
were left with 6 additional studies
\citep{Park2006, Lowsky2013, Lee2015, Ng2015, Lee2017, Wang2019}, 1 of
which was excluded from the synthesis for using \hl{kNN} prediction. Reference
tracking from the 43 + 360 = 403 studies assessed for eligibility led us
to identify 1 additional study that met our criteria
\citep{Kasabov2010}. This left us with 21 + 1 + 5 = 27 studies included
in the review and synthesis.

@@

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While these studies trace back to the late 1990s, they
have become \hl{slightly }more common, which suggests that this is an
active, though not rapidly expanding, approach. We note that CBR in
health and medicine originated as early as 1990, and \hl{SCBM} emerged
soon after CBR had established itself; the idea has been ``in the air''
for as long as CBR has been in use.\hl{ Based on the motivations and methodologies observed in our sample (discussed in later sections), we speculate that the higher frequency of studies since 2013 is attributable in large part to the availability of vaster clinical databases with improved pre-processing tools, following the rapid uptake of EHRs; and of more efficient and powerful tools for retrieval (similarity-matching), the traditional bottleneck of CBR. Another possibility is an increased interest in interpretable machine learning }\citep{Rudin2022}\hl{. Each of these factors should apply with no less force to other CDSS and clinical applications of ML, and the rise we see may simply be commensurate with an overall increase in the volume of articles in these fields.}

@@

While all included studies were reported in scientific and medical
journals, the vast majority were oriented toward clinical practice
rather than medical research. For example, a 2006 study specifically
evaluated the usefulness of CBR-based explanations for the purpose of
decision support \citep{Doyle2006}. This study was part of a much larger
literature on CBR systems and was included here despite relying on \hl{kNN}
prediction because it used an unconventional voting scheme to generate
recommendations. A more recent study took essentially the same focus
with respect to a proposed clinical risk prediction model, which
amounted to CBR with a novel weighting scheme on predictors informed by
expert consensus \citep{Fang2021}. In both cases a prototype
implementation was \hl{evaluated} in an experimental \hl{setting.}

@@

\hl{Eight} studies used response values in training data sets to
supervise similarity learning while two used unsupervised learning
(Table \ref{tab:composite}). Based on the key properties of DML
algorithms identified by \citet{Bellet2014}, most learned measures were
non-linear while some were locally learned, and some were optimized
globally while others locally. Not all qualified as DML, since the
resulting measures would not necessarily satisfy the triangle
inequality. Each used its similarity measure to retrieve training cases
relevant to each testing case. \hl{Most} took a standard \hl{kNN} approach to
generating predictions; \hl{two} \citep{Vilhena2016, Tang2021} fit
predictive models to the retrieved cohorts.

@@

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by \citet{Park2006} to
optimize the predictive performance of models fitted to cohorts
retrieved using a fixed similarity rather than count threshold, a
counterpart to conventional CBR they called ``statistical
CBR''.\footnote{Drawing from \citet{Goyal2008}, we suggest the term
  ``combinatorial CBR'' for the then-conventional approach using
  similarity cohorts of fixed cardinality.} The step used a heuristic
procedure to locate a similarity threshold that (locally) maximizes
predictive accuracy on the training set. This is analogous to optimizing
the neighborhood size parameter in \hl{kNN} predictive modeling, so we will
not discuss it further except to mention that the authors found
statistical CBR to outperform conventional CBR on several data sets.
\citet{CampilloGimenez2013} combined these learning techniques: They
used logistic regression on the training set to obtain weights for the
predictors (by predicting the binary outcome of kidney transplant
waitlist registration) and for the cases (by predicting agreement of
outcome between an index case and other training cases), and they used
exhaustion to optimize the size of the retrieved cohort for the accuracy
of the \hl{kNN} model using the previously optimized weights.

@@

\hl{One theme} was the optimization of retrieved cohort sizes for some
measure of model performance. Despite the early recommendation by
\citet{Park2006} to threshold cohorts by similarity rather than by
cardinality, only the one previously published study
\citep{Mariuzzi1997} and the most recent study \citep{Tang2021, Ng2021}
in this corpus took this approach, though \citet{Liu2022} took a third
option by retrieving fixed proportions of training data. Viewed as a
hyperparameter of the individualized modeling approach, cohort size can
be treated in the same way as neighborhood size in \hl{kNN} \hl{prediction},
so we will not discuss it further.

@@

\hl{While most studies identified outstanding technical needs (Section }\ref{identified-needs}\hl{), few (though some) emphasized the need to assess interpretability, meaningfulness, or trust.
Most studies identified interpretability as an advantage of localized models, though none evaluated model interpretability and few proposed new interpretative uses.
Lack of trust in ML tools is a long-recognized problem that direct intepretability of model components could help alleviate.
One valuable direction for future work would be to measure the utility of interpretable components in research and in clinical practice and the correctness and confidence of users in their interpretations.
Another would be to compare the localized predictor importance measures obtained from localized models to the local importance measures used to explain predictions made by opaque models.}

@@

\begin{itemize}
\tightlist
\item
  CBR: case-based reasoning
\item
  CIS: clinical information system
\item
  \hl{CDSS:} clinical decision support\hl{ system}
\item
  AI: artificial intelligence
\item
  \hl{kNN}: nearest neighbors
\item
  GLM: generalized linear model
\item
  \hl{DT: decision tree}
\item
  \hl{RF: random forest}
\item
  \hl{ANN: artificial neural network}
\item
  ML: machine learning
\item
  PMC: PubMed Central
\item
  LOOCV: leave-one-out cross-validation
\item
  GSA: gravitational search algorithm
\item
  DML: distance metric learning
\end{itemize}

@@

& & & & \hspace{6em} & \hl{kNN} & 88.29\%\hspace{6em} \\

@@

& & & & \hspace{6em} & \hl{kNN} & 82.22\%\hspace{6em} \\

@@

& & Breast & Statistical CBR & 96.96\%\hspace{6em} & \hl{kNN} &

@@

& & & & \hspace{6em} & \hl{kNN} & 65.39\%\hspace{6em} \\

@@

& & & & \hspace{6em} & \hl{kNN} & 55.59\%\hspace{6em} \\

@@

\citet{Malykh2018} & Accuracy & J13/pneumonia & Case-based \hl{ANN} &

@@

& & K80.1/gallbladder & Case-based \hl{ANN} & 76.7\%\hspace{6em} & &

@@

& & H25.1/age-related cataract & Case-based \hl{ANN} & 94.9\%\hspace{6em} & &

@@

& & H26.2/complicated cataract & Case-based \hl{ANN} & 91.4\%\hspace{6em} & &

@@

& & 167.4/encephalopathy & Case-based \hl{ANN} & 72.4\%\hspace{6em} & &

@@

& & 167.9/CBD & Case-based \hl{ANN} & 75.4\%\hspace{6em} & & \hspace{6em} \\
& & N20.1/ureter & Case-based \hl{ANN} & 58.7\%\hspace{6em} & &
